---
title: "Export UK Office of National Statistics Lifetables"
author: "Altan Orhon"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: rmarkdown::github_document
params:
  data_url: "https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/birthsdeathsandmarriages/lifeexpectancies/datasets/nationallifetablesunitedkingdomreferencetables/current/nltuk198020203.xlsx"
---

This notebook downloads UK Office of National Statistics Lifetables for multiple years, turns them into long format, and exports them to `TSV` and `RDS`.

```{r setup, include=FALSE}
if (!requireNamespace("rlang", quietly = TRUE)) install.packages("rlang")
requireNamespace("rlang", quietly = TRUE)
rlang::check_installed(c("knitr", "tidyr", "purrr", "dplyr", "cellranger", "readxl", "withr", "checkmate", "here"))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Download data:

```{r download-data}
data_path <- params$data_path %||% here::here("data", basename(params$data_url))
dir.create(data_path, showWarnings = FALSE, recursive = TRUE)
if (!file.exists(data_path)) {
  message("Downloading data to ", data_path)
  download.file(data_url, data_path)
}
```

Function to load dataset from the `XLSX` file:

```{r nltuk-functions}
load_nltuk_dataset <- function(path = "data/nltuk198020203.xlsx", sheet_names = NULL, sheet_pattern = "(^[2][0-9]{3})-([2][0-9]{3})", sheet_range = cellranger::cell_limits(c(6, 1), c(NA, 6)), year_min = -Inf, year_max = Inf, ...) {
  if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
  requireNamespace("readxl", quietly = TRUE) #
  if (is.null(sheet_names)) {
    sheet_names <- readxl::excel_sheets(path) |>
      purrr::keep(stringr::str_detect, sheet_pattern) |>
      unique()
    checkmate::assert_character(sheet_names, min.len = 1)
    cli::cli_inform("Identified {length(sheet_names)} sheet names ({sheet_names})")
  }
  if (is.null(names(sheet_names))) {
    names(sheet_names) <- as.character(sheet_names)
  }
  if (inherits(sheet_range, "cell_limits")) {
    sheet_range <- rep(list(sheet_range), length(sheet_names)) |> rlang::set_names(sheet_names)
  }
  checkmate::assert_list(sheet_range, len = length(sheet_names))
  res <- purrr::map2(sheet_names, sheet_range, ~ {
    readxl::read_excel(path,
      sheet = .x,
      range = .y,
      col_types = c(age = "numeric", mx = "numeric", qx = "numeric", lx = "numeric", dx = "numeric", ex = "numeric")
    )
  }) |>
    dplyr::bind_rows(.id = "sheet") |>
    dplyr::arrange(dplyr::desc(sheet), age) |>
    tidyr::separate_wider_delim(sheet, delim = "-", names = c("y1", "y2"), cols_remove = FALSE) |>
    dplyr::mutate(y1 = as.integer(y1), y2 = as.integer(y2))

  res |>
    dplyr::group_by(sheet) |>
    dplyr::reframe(year = min(y1):max(y2)) |>
    dplyr::ungroup() |>
    dplyr::filter(year >= year_min, year <= year_max) |>
    dplyr::left_join(res, by = "sheet", relationship = "many-to-many") |>
    dplyr::group_by(year, age) |>
    dplyr::arrange(dplyr::desc(year)) |>
    # Use the most recent value for that year
    dplyr::filter(dplyr::row_number() == 1) |>
    dplyr::ungroup() |>
    dplyr::arrange(year, age) |>
    dplyr::select(-y1, -y2) |>
    dplyr::rename(origin = sheet)
}
```

Load the data:

```{r load-data}
# Load everything from 2004
nltuk_df <- load_nltuk_dataset(data_path, year_min = 2004)
```

## Data Description

- `mx` is the central rate of mortality (number of deaths at age `x` last
birthday in the 3-year period to which the life table relates divided by
the average population at that age over the same period)

- `qx` is the mortality rate between age `x` and `x + 1` (probability that
a person aged `x` exactly will die before reaching age (`x` +1))

- `lx` is the number of survivors to exact age `x` of 100,000 live births
of the same sex (who are assumed to be subject throughout their lives to
the mortality rates experienced in the 3-year period to which the life
table relates)

- `dx` is the number dying between exact age `x` and `x + 1`, described
similarly to `l[x]`, that is `d[x] = l[x] - l[x +1]`.

- `ex` is the average period expectation of life at exact age `x`, that
is the average number of years that those aged `x` exactly will live
thereafter based on the mortality rates experienced in the 3-year period
to which the life table relates.

## Preview data

```{r}
head(nltuk_df)
```

## Export data

Export TSV:

```{r export-tsv}
output_path_tsv <- params$output_path_tsv %||% here::here("data", paste0(tools::file_path_sans_ext(basename(data_path)), "_", paste0(range(nltuk_df$year, na.rm = TRUE), collapse = "-"), ".tsv"))
message("Exporting TSV to ", output_path_tsv)
dir.create(dirname(output_path_tsv), recursive = TRUE, showWarnings = FALSE)
if (!file.exists(output_path_tsv)) {
  message("Writing data to ", output_path_tsv)
  write.table(nltuk_df, sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE, file = output_path_tsv)
} else {
  message("Data already exists at ", output_path_tsv)
}
```

Export RDS:

```{r export-rds}
output_path_rds <- params$output_path_rds %||% here::here("data", paste0(tools::file_path_sans_ext(basename(data_path)), "_", paste0(range(nltuk_df$year, na.rm = TRUE), collapse = "-"), ".rds"))
message("Exporting RDS to ", output_path_rds)
dir.create(dirname(output_path_rds), recursive = TRUE, showWarnings = FALSE)
if (!file.exists(output_path_rds)) {
  message("Writing data to ", output_path_rds)
  saveRDS(nltuk_df, output_path_rds)
} else {
  message("Data already exists at ", output_path_rds)
}
```

## Session Info

```{r}
sessionInfo()
```
